Continuous Deductive Reasoning over Streamed RDF Data

### 2015-07-13 content/contribution/continuous-datalog.tex
Various changes

In addition, it can be shown that these streams may be trivially
manipulated to retrieve the instantaneous sets of entailments produced
by the more semantically well-defined offerings from the reasoning
community.

@@

We define the concept of entailment in Datalog as follows. A fact $e$
is entailed by a program $\Pi$ iff each interpretation satisfying
$\Pi$ (satisfying every clause in $\Pi$) also satisfies $e$, written
$\Pi \models e$. For a a program $\Pi$ and a set of entailed facts
$E$, $\Pi \models E$ iff $\Pi \models e$ for every $e \in E$. For
convenience, we refer to the set of all facts that are entailed by a
program (the intersection of all the Herbrand models of the program)
as the {\em entailment of a program}, written $cons(\Pi)$ and defined
as:
\begin{align*}
  cons(\Pi) &= \{ e \in \hat{H} \mid \Pi \models e \} \\
  &= \bigcap_\mathcal{I} \models_\mathcal{I} \Pi
\end{align*}
\end{definition}

@@

When considering a TA-Datalog program, in which there may
be many distinct instances of a given fact, temporal information
regarding when instances of that fact became true would be lost if
such a program were to entail sets of plain facts without temporal
annotations. We therefore extend the model theory for Datalog from
\refdef{continuous datalog: datalog program} by considering the
Herbrand base of all fact instances $\hat{H}^T$ (all facts at all
instants). As before, a fact instance $e$ is entailed by a program
$\Pi^T$ iff each interpretation satisfying every clause in $\Pi^T$
also satisfies $e$. The entailment of the program $\Pi^T$ is therefore:
\[ cons^T(\Pi^T) = \{ e \in \hat{H}^T \mid \Pi^T \models e \} \]

@@

Consider the following example TA-Datalog program:
\begin{align*}
  \Pi^T = \langle {} & \{ a \Leftarrow b \land c \}, \\
  & \{ \langle a, 1 \rangle, \langle a, 2 \rangle, \langle b, 1 \rangle, \langle b, 2 \rangle  \} \rangle
\end{align*}

The entailment of the program $\Pi^T$ will be:

\[cons^T(\Pi^T) = \{\langle c, 1 \rangle, \langle c, 2 \rangle\}\]

It follows that applying the ${strip()}$ function to all the fact
instances in both the set of fact instances $F^T$ in some ${\Pi^T =
  \langle R, F^T \rangle}$ and the entailment of the program $\Pi^T$
(the set of fact instances $cons^T(\Pi^T)$) maintains the relationship
of entailment; if $\langle R, F^T\rangle \models \langle e, t
\rangle$, then $\langle R, strip(F^T) \rangle \models strip(\langle e,
t \rangle)$.

This relationship is shown in \reffig{TD-D}.

@@

Reconsidering the example TA-Datalog program above, we may convert it
into the following equivalent conventional Datalog program:
\[
  \Pi = \langle  \{ a \Leftarrow b \land c \}, \{ a, b \} \rangle
\]
Trivially, the entailment of this program $\Pi$ will therefore be:
\[cons(\Pi) = \{ c \}\]
\end{definition}

@@

Having defined the notion of time annotation in Datalog, we now extend
this concept to that of Continuous Datalog (henceforth C-Datalog).

\begin{definition}[Streams]
\labeldef{continuous datalog: stream}
A {\em stream} is a sequence of fact instances ordered
by their instants; in conjunction with the distinction between
multiple instances of a fact by their timestamp, this definition justifies
Axiom~\ref{axiom:continuous datalog: window disjointness}.
\end{definition}

\begin{definition}[Sliding Window]
\labeldef{continuous datalog: window}

A sliding window of range ${x \in \mathbb{N}}$ over a stream $S$ at
an instant ${i \in \mathbb{T}}$ is the \emph{set} of fact
instances that occur in $S$ during the interval ${(i-x,i]}$, written
${W^{S,x}_{i}}$, and defined as follows:

\[ W^{S,x}_i = \{ \langle f, t \rangle \mid \langle f, t \rangle \in S \land (i-x) < t \leqslant i \}\]

@@

From this description of instance entailment,
Axiom~\ref{axiom:continuous datalog: entailment disjointness} may be
derived:

\begin{axiom}\label{axiom:continuous datalog: entailment disjointness}
If two sliding windows over the same stream $S$ cover non-overlapping time periods,
then a set ${E^T1}$ of fact instances entailed by ${\langle R, W^{S,x}_i \rangle}$
is necessarily disjoint from any set ${E^T2}$ of fact instances entailed by
${\langle R, W^{S,y}_{i-z} \rangle}$, where ${z \geq x}$.
% This is because the timestamps of the instances ${E^T1}$ are a subset of those of
% the instances $W^{S,x}_i$, the timestamps of the instances ${E^T2}$ are a subset
% of those of the instances $W^{S,y}_{i-z}$, fact instances are distinct by timestamp and
% ${\{ t : f | t \in W^{S,x}_i\} \cap \{ t : f | t \in W^{S,y}_{i-z}\} = \varnothing}$.
\[ \{ e : \langle R , W^{S,x}_i \rangle \vdash e \} \cap \{ e : \langle R , W^{S,y}_{i-z} \rangle \vdash e \} = \varnothing \]
\end{axiom}

@@

\begin{definition}[Continuous Datalog Program]\label{def:continuous datalog: CDP}
A Continuous Datalog Program $\Pi^C$ is a tuple ${\langle R, F, S, r
\rangle}$ which comprises a finite set of rules $R$, a finite set of persistent
facts $F$, a stream of facts $S$ and a range ${r \in \mathbb{N}}$ for
the sliding window over $S$. This specifies that the streamed instances from $S$ are
to be considered valid from their timestamp for the duration represented
by $r$.

The entailment of such a program is composed of a stream of entailed
instances $P$ and a stream of instance negations $N$, whose contents
are defined in 
Axioms~\ref{axiom:continuous datalog: positive window increment}, 
\ref{axiom:continuous datalog: negative window increment} and
\ref{axiom:continuous datalog: entailment precedes negation}. 
These specify that the streams $P$ and $N$ are the continuous
reasoning equivalents of the streams produced by the \emph{IStream}
and \emph{DStream} operators of the \emph{CQL} continuous query
language \citep{CQL}, respectively.
\[ \Pi^C = \langle R, F, S, r \rangle \vDash \langle E, P, N \rangle \]
The truth value of a $\Pi^C$ is instantaneous, with the the truth
function of Continuous Datalog $T^{CD}$ taking both a program ${\Pi^C}$
and an instant ${i \in \mathbb{N}}$ as arguments.
\end{definition}

>>>>

In addition, we show that these streams may be manipulated to retrieve
the instantaneous sets of entailments produced by the more
semantically well-defined offerings from the reasoning community.

%%% what does this mean? ^

@@

We define the concept of entailment in Datalog as follows. A fact $e$
is entailed by a program $\Pi$ iff each interpretation satisfying
$\Pi$ (satisfying every clause in $\Pi$) also satisfies $e$, written
$\Pi \models e$. For a program $\Pi$ and a set of entailed facts
$E$, $\Pi \models E$ iff $\Pi \models e$ for every $e \in E$. For
convenience, we refer to the set of all facts that are entailed by a
program (the intersection of all the Herbrand models of the program)
as the {\em entailment of a program}, written $cons(\Pi)$ and defined
as:
\begin{align*}
  cons(\Pi) &= \{ e \in \hat{H} \mid \Pi \models e \} \\
  &= \bigcap_\mathcal{I} \models_\mathcal{I} \Pi
\end{align*}
\end{definition}

@@

When considering a TA-Datalog program, in which there may be many
distinct instances of a given fact, temporal information regarding
when instances of that fact became true would be lost if such a
program were to entail sets of plain facts without temporal
annotations. We therefore extend the model theory for Datalog from
\refdef{continuous datalog: datalog program} by considering the
Herbrand base of all fact instances $\hat{H}^T$ (all facts at all
instants). As before, a fact instance $e$ is entailed by a program
$\Pi^T$ iff each interpretation satisfying every clause in $\Pi^T$
also satisfies $e$, written $\Pi^T \models e$. The {\em entailment of the
program} $\Pi^T$ is therefore:
\[ cons^T(\Pi^T) = \{ e \in \hat{H}^T \mid \Pi^T \models e \} \]

@@

Consider the following example TA-Datalog program:
\[
\Pi^T = \langle \{ c \Leftarrow a \land b \}, \{ \langle a, 1 \rangle, \langle a, 2 \rangle, \langle b, 1 \rangle, \langle b, 2 \rangle  \} \rangle
\]

The entailment of the program $\Pi^T$ will be:
\[cons^T(\Pi^T) = \{\langle c, 1 \rangle, \langle c, 2 \rangle\}\]

It follows that applying the ${strip()}$ function to all the fact
instances in both the set of fact instances $F^T$ in some ${\Pi^T =
  \langle R, F^T \rangle}$ and the entailment of the program $\Pi^T$
(the set of fact instances $cons^T(\Pi^T)$) maintains the relationship
of entailment; if $\langle R, F^T\rangle \models \langle e, t
\rangle$, then $\langle R, strip(F^T) \rangle \models strip(\langle e,
t \rangle)$. This relationship is shown in \reffig{TD-D}.

@@

Reconsidering the example TA-Datalog program above, we may convert it
into the following equivalent conventional Datalog program $\Pi$:
\begin{align*}
\Pi &= \langle \{ c \Leftarrow a \land b \}, strip(\{ \langle a, 1 \rangle, \langle a, 2 \rangle, \langle b, 1 \rangle, \langle b, 2 \rangle \}) \rangle \\
    &= \langle \{ c \Leftarrow a \land b \}, \{ a, b \} \rangle
\end{align*}
Trivially, the entailment of this program $\Pi$ will therefore be:
\begin{align*}
cons(\Pi) &= strip(\{\langle c, 1 \rangle, \langle c, 2 \rangle\}) \\
 &= \{ c \}  
\end{align*}
\end{definition}

@@

Having defined the notion of time annotation in Datalog, we now define
the notions of streams and windows before extending to Continuous Datalog.

\begin{definition}[Streams]
\labeldef{continuous datalog: stream}
A {\em stream} is a sequence of fact instances ordered
by their instants.
%FIXME: does this mean that a stream never contains duplicates?
%; in conjunction with the distinction between
%multiple instances of a fact by their timestamp, this definition justifies
%Axiom~\ref{axiom:continuous datalog: window disjointness}.
\end{definition}

\begin{definition}[Window]
\labeldef{continuous datalog: window}

A window of range ${x \in \mathbb{N}}$ over a stream $S$ at an instant
${i \in \mathbb{T}}$ is the \emph{set} of fact instances that occur in
$S$ during the interval ${(i-x,i]}$, written ${W^{S,x}_{i}}$, and
defined as follows:
\[ W^{S,x}_i = \{ \langle f, t \rangle \mid \langle f, t \rangle \in S \land (i-x) < t \leqslant i \}\]  
\end{definition}

@@

From this description of instance entailment,
Axiom~\ref{axiom:continuous datalog: entailment disjointness} may be
derived:
\begin{axiom}\label{axiom:continuous datalog: entailment disjointness}
  If two sliding windows over the same stream $S$ cover non-overlapping time periods,
then a set $E^T_1$ of fact instances entailed by ${\langle R, W^{S,x}_i \rangle}$
is necessarily disjoint from any set $E^T_2$ of fact instances entailed by
$\langle R, W^{S,y}_{i-z} \rangle$, where ${z \geq x}$.
% This is because the timestamps of the instances ${E^T1}$ are a subset of those of
% the instances $W^{S,x}_i$, the timestamps of the instances ${E^T2}$ are a subset
% of those of the instances $W^{S,y}_{i-z}$, fact instances are distinct by timestamp and
% ${\{ t : f | t \in W^{S,x}_i\} \cap \{ t : f | t \in W^{S,y}_{i-z}\} = \varnothing}$.
\[ \{ e : \langle R , W^{S,x}_i \rangle \vdash e \} \cap \{ e : \langle R , W^{S,y}_{i-z} \rangle \vdash e \} = \varnothing \]
\end{axiom}

@@

\begin{definition}[Continuous Datalog Program]
\labeldef{continuous datalog: CDP}

A Continuous Datalog program (henceforth abbreviated C-Datalog)
$\Pi^C = \langle R, F^0, S, r \rangle$ is a tuple which comprises a
finite set of rules $R$, a finite set of static fact instances $F^0$
all of which have the instant $0$, a stream $S$ and a range ${r \in
\mathbb{N}}$ for the window over $S$.

%This specifies that the streamed instances from $S$
%are to be considered valid from their timestamp for the duration
%represented by $r$.

We consider the entailment of the C-Datalog program $\Pi^C$ not just
as a set of fact instances (as for TA-Datalog), but in three parts: a
set of fact instances $E^0$ entailed by the fact instances in $F^0$, a
stream $P$ of fact instances entailed by the fact instances in $F^0$
and S, and a stream $N$ of fact instance negations. $P$ and $N$ can be
considered as the continuous reasoning equivalents of the streams
produced by the \emph{IStream} and \emph{DStream} operators of the
\emph{CQL} continuous query language \citep{CQL}, respectively.

whose contents
are defined in 
Axioms~\ref{axiom:continuous datalog: positive window increment}, 
\ref{axiom:continuous datalog: negative window increment} and
\ref{axiom:continuous datalog: entailment precedes negation}. 
These specify that the streams 
\[ \Pi^C = \langle R, F, S, r \rangle \vDash \langle E, P, N \rangle \]
The truth value of a $\Pi^C$ is instantaneous, with the the truth
function of Continuous Datalog $T^{CD}$ taking both a program ${\Pi^C}$
and an instant ${i \in \mathbb{N}}$ as arguments.
\end{definition}
